---
title: "Factor Analysis of World Universities"
author: "Jeremy Leipzig"
date: "April 11, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(psy)
library(GGally)
library(stringr)
library(gplots)
```
## Data
I am using data about world university rankings provided by Kaggle:

[https://www.kaggle.com/mylesoneill/world-university-rankings](https://www.kaggle.com/mylesoneill/world-university-rankings)

I used two tables in this dataset
* Times Higher Education World University Ranking 
* Center for World University Rankings

### Times Fields:
* world_rank - world rank for the university. Contains rank ranges and equal ranks (eg. =94 and 201-250).
* university_name - name of university.
* country - country of each university.
* teaching - university score for teaching (the learning environment).
* international - university score international outlook (staff, students, research).
* research - university score for research (volume, income and reputation).
* citations - university score for citations (research influence).
* income - university score for industry income (knowledge transfer).
* total_score - total score for university, used to determine rank.
* num_students - number of students at the university.
* student_staff_ratio - Number of students divided by number of staff.
* international_students - Percentage of students who are international.
* female_male_ratio - Female student to Male student ratio.
* year - year of the ranking (2011 to 2016 included).

### CWUR Fields:
* world_rank - world rank for university.
* university_name - name of university.
* country - country of each university.
* national_rank - rank of university within its country.
* quality_of_education - rank for quality of education.
* alumni_employment - rank for alumni employment.
* quality_of_faculty - rank for quality of faculty.
* publications - rank for publications.
* influence - rank for influence.
* citations - rank for citations.
* broad_impact - rank for broad impact (only available for 2014 and 2015)
* patents - rank for patents.
* score - total score, used for determining world rank.
* year - year of ranking (2012 to 2015).

```{r loadit}
read.csv("world-university-ranking/cwurData.csv",header=TRUE,stringsAsFactors = FALSE) %>% dplyr::filter(year==2015 | year==2014) %>% select(-country) -> cwur
read.csv("world-university-ranking/timesData.csv",header=TRUE,stringsAsFactors = FALSE) %>% dplyr::filter(year==2015 | year==2014) %>% select(-world_rank)  -> times
names(cwur)<-paste("cw",names(cwur),sep="_")
names(times)<-paste("ti",names(times),sep="_")
merge(x=times,y=cwur,by.x = c("ti_university_name","ti_year"),by.y=c("cw_institution","cw_year"),suffixes = c(".times",".cwur"),all=FALSE) %>% arrange(ti_university_name,-ti_year) %>% distinct(ti_university_name) -> univ
knitr::kable(t(head(univ)))
```

```{r munge}
univ$ti_international<-as.numeric(univ$ti_international)
univ$ti_income<-as.numeric(univ$ti_income)
univ$ti_female_male_ratio<-as.numeric(str_extract(univ$ti_female_male_ratio,'\\d+'))
univ$ti_num_students<-as.numeric(gsub(",", "", as.character(univ$ti_num_students)))
univ$ti_international_students<-as.numeric(gsub("%", "", as.character(univ$ti_international_students)))
nums <- sapply(univ, is.numeric)
nums[1] <- TRUE
nums[3] <- TRUE
univ[,nums] %>% select(-ti_year,-ti_female_male_ratio) %>% na.omit -> univ_na
nums[1] <- FALSE
nums[3] <- FALSE
univ[,nums] %>% select(-ti_year,-ti_female_male_ratio) %>% na.omit -> univ_df
row.names(univ_df) <- univ_na$ti_university_name

# if you want to convert ti scores to ordinals, you can do it here
# but it actually decreases the loading of those variables
univ_ranked<-cbind(univ_df[,c(9:19)],lapply(univ_df[,c(1:8)],function(x){order(x,decreasing=TRUE)}))

univ_df %>% as.matrix -> univ_num
```

## Heatmap and dendrogram
```{r dendro, fig.width=10, fig.height=40}
heatmap.2(as.matrix(scale(univ_num)),margins = c(10, 10),trace="none",cexRow=0.68, cexCol=0.9)
```

## PCA
Let's first run a PCA to get a rough idea of the dimensionality of the data. A PCA should help to identify factors prior to factor analysis.
```{r pca}
pca.univ<-princomp(univ_num, corr=TRUE)
summary(pca.univ)
```

So the first component provides 99% of the variance.

## Number of Factors to Retain

###Scree plot
```{r scree}
plot(pca.univ,type="lines")
```

To be honest this looks like one component to rule them all, but...

```{r fa}
fa<-factanal(univ_num,factors=3,rotation="none")
fa
```

OK so the first component isn't everything - just 39% of the variance.

## Rotate
```{r farotate}
fa_rotate<- factanal(univ_num,factors=3,scores="regression",rotation="varimax")
fa_rotate
```

Actually the 25%, 21%, and 10% of the variance are explained for the first three factors respectively.

## Compare
```{r comprot}
fa_load<-as.data.frame(unclass(loadings(fa)))
fa_rotate_load<-as.data.frame(unclass(loadings(fa_rotate)))
names(fa_rotate_load)<-paste("rot",names(fa_rotate_load),sep="_")
knitr::kable(cbind(fa_load,fa_rotate_load))
```

## Interpret
```{r interpret}
bolden<-function(x){if(abs(x[1])>=.40){return(paste("**",x,"**",sep=""))}else{return(as.character(x))}}
bolden_fact<-sapply(fa_rotate_load,bolden)
fa_rotate_load$factor1<-sapply(fa_rotate_load$rot_Factor1,bolden)
fa_rotate_load$factor2<-sapply(fa_rotate_load$rot_Factor2,bolden)
fa_rotate_load$factor3<-sapply(fa_rotate_load$rot_Factor3,bolden)
bolden_fact<-fa_rotate_load[,c("factor1","factor2","factor3")]
knitr::kable(bolden_fact)
```
CW-derived ranks appear to be more in line with the first two factors

## US Schools
Let's look at the subset of 70 US schools to help interpret these factors

```{r usschools, fig.width=15, fig.height=15}
scores<-as.data.frame(fa_rotate$scores)
scores$country<-univ_na$ti_country
scores$university<-univ_na$ti_university_name
scores %>% filter(country=='United States of America') -> usschools
ggplot(usschools,aes(Factor1,Factor2))+geom_point()+geom_text(aes(label=university),hjust=0, vjust=0)
```

* Factor 1 seems to be related to humanities/engineering focus
* Factor 2 is in line with general prestige or competitiveness with a favor toward larger graduate programs
* Factor 3 is a measure of the international or provincial nature of these schools
